Using the species abundance distribution to validate ecological and palaeoecological data

### 2024-03-01 count_check_MS.qmd
update renv
new intro

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so, and perhaps other aspects of the counts. This paper explores the statistical properties of ecological and palaeoecological species counts and uses the findings to develop methods to flag count or percentage data with unexpected distributions. Several assemblage datasets that fail these tests, and potentially have misreported counts or count sums, are identified.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: manuscript_extra/countMS2.bib
csl: manuscript_extra/elsevier-harvard_rjt.csl
execute: 
  echo: false
  error: false
---

@@

# tar_load("summarised_bird_singletons")
# tar_load("summarised_bird")
# tar_load("testate_summ")
 tar_load("last_chance_summ")
# tar_load("pollen_assem_res")
# tar_load("pollen_dataset_res")
 tar_load("pollen_results")
 tar_load("hershop_pollen")
 tar_load("wildhorse")
 tar_load("pollen3")
 tar_load("pollen1")
 tar_load("pollen2")
# tar_load("owens_summ1")
# tar_load("owens_summ")
# tar_load("owens_est_summ")
 tar_load("diatom1_summ")
 tar_load("diatom2_summ")
 tar_load("chironomid1_summ")
 tar_load("marine1_summ")
# tar_load("chironomid2_summ")

@@

Six percent of papers published in _Molecular and Cellular Biology_ show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct.
It would, especially after the spate of retractions in behavioural ecology [@Pennisi2021], be reckless to assume that the ecological and palaeoecological literature is devoid of errors and misconduct. 
The low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence.

Several numerical tools have been developed to identify questionable data.
Deviations from the distribution of digits expected from the Newcomb-Benford law have been used to detect issues with scientific and financial data [@Barabesi2017].
@Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data.
@Brown2017 developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size.
However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data.

A common assertion in papers reporting microfossil assemblage data is that a minimum of _N_ microfossils were counted, where _N_ is often fifty for chironomids and several hundred for pollen and diatoms.
This is important because larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
There is a risk that count sums will be misreported.
This could happen by mistake, especially with the use of secondary data.
It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low.
This risk that count sums could be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen.
@Telford2019 reports some other cases where count sums may be much lower than reported.

Given that count sums have been misreported, the possibility that other aspects of count data may be misreported needs to be considered.

The key insight that allows inference about community or assemblage count data is that they can be represented by a species abundance distribution.
The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic curve with many rare and few abundant species and is central to ecological theory [@McGill_2007].
From this, two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, and that the distribution of abundances should have a smooth and monotonic decline.
Count data that differ markedly from these predictions might not accurately reflect the original community or assemblage.

This paper aims to develop simple methods to test the abundance of rare taxa and the smoothness of the species abundance distribution that can be used to flag cases where the count data might have been misreported.
Next, as, regrettably, many palaeoecological data sets are archived as percent data, these tests are then extended to identify percent data where the count sum might be lower than reported.
Finally, some complications and caveats are discussed.

@@

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness.

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along ~40 km survey routes at the height of the breeding season. 
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database [@neotoma_2018] downloaded on `r targets::tar_meta(pollen_data) |> pull(time) |> format("%d %B %Y")`.

Counts with with small count sums (< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded from all datasets.

@@

Data sets with a many samples without singletons, a greatest common divisor greater than one, or an unusually high proportion, relative to the majority of pollen data set in the Neotoma database, of counts divisible by $k$ were flagged for closer inspection.

@@

In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given the expectation that the rarest taxon is a singleton, the count sum _N_ can be estimated as $1/p_{min} \times 100$ where _p~min~_ is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
It is also expected that the percent _p_ is calculated from integer counts, therefore there should be a count sum _N_ such that $p_i/100\times N$ is, within rounding error, an integer for all _i_ taxa.
Possible values for _N_ can be found by a direct search over the range of plausible values of _N_.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of _N_ except in cases where the greatest common divisor is greater than one.
These tests for percent data are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

All analyses were done in R v. `r getRversion()` [@R] and used the packages 
extraDistr v. `r packageVersion("extraDistr")` [@extraDistr],
numbers v. `r packageVersion("numbers")` [@numbers], 
countSum v. `r packageVersion("countSum")` [@countSum], 
neotoma2 v. `r packageVersion("neotoma2")` [@neotoma2], 
bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], 
and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Packages targets v. `r packageVersion("targets")` [@targets], 
quarto v. `r packageVersion("quarto")` [@bookdown1] and
renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
Code to reproduce all the analyses is archived at [https://github.com/richardjtelford/count.check.ms](https://github.com/richardjtelford/count.check.ms) and [zenodo.org](zenodo.org).
All issues identified in the Neotoma database were reported to the maintainers.

@@

Some datasets with assemblages without singletons have few taxa.
For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@Mehringer1985] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of Neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`). 
While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts. 
This is in contrast to most (`r pollen_summ$has_zeros * 100`%) Neotoma pollen assemblages which have zero counts.  
These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.

@@

`r Words(pollen1$n_gcd2)` of the `r pollen1$n_sample` assemblages in dataset pollen1 lack singletons and have a greatest common divisor of two.
These assemblages have between `r pollen1$min_taxa` and `r pollen1$max_taxa` taxa.
The probability of all _n_ counts being divisible by _k_ is $k^{-n}$, assuming the distribution of counts is locally uniform.
That is, for _k_ = 2, odd counts are as common as even counts.
For an assemblage with `r pollen1$min_taxa` or `r pollen1$max_taxa` taxa, the probability of having all counts divisible by two by chance are 10^`r ceiling(- pollen1$min_taxa * log10(2))`^ and 10^`r ceiling(- pollen1$max_taxa * log10(2))`^ respectively.
These probabilities are not so small that it is unrealistic to expect to find assemblages like these in the Neotoma database, but to find so many in one dataset is unexpected.
Furthermore, the assumption made here is optimistic, in reality for _k_ = 2, more odd counts are expected ones are expected than even ones in most datasets, as typical rank abundance curves have more singletons than doubletons and so on, which makes finding assemblages with a greatest common divisor above one even more unlikely.

Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (@fig-multiples-plot).
It is unclear what processes could generate such counts.

@@

## Unexpected percent data
```{r}
#| label: line169

@@

Some taxonomic groups have microfossils that often come in groups of attached individuals.
For example, each diatom cell has two valves which are the counting unit.
These valves usually separate during processing, but paired valves are often found, and some taxa such as _Aulacoseira_ produce long chains of strongly bound valves.
This might make singletons slightly less likely. 
The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.

@@

# Acknowledgements{-}

@@

# References{-}

>>>>

---
title: "Using the species abundance distribution to validate ecological and palaeoecological data"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |

  Species community or assemblage counts are a key data type in ecology and palaeoecology. 
  Sometimes, the data are archived as percentages.
  Unfortunately, necessary metadata such as the count sum are not always archived with the percentages.
  This paper uses insights from the ubiquitous shape of specied abundance distributions to develop methods to estimate the count sum of percentage data when it is missing.
  The methods are demonstrated on a collection of archived percent data.   
  The assumptions of the methods are tested using a large collection of ecological and palaeoecological count data.
  Several datasets that fail the tests. 
  These potentially have misreported data.
  The consequences of misreported data for different types of analyses is discussed.
date: today
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
format: pdf
documentclass: article
classoption: a4paper
bibliography: manuscript_extra/countMS2.bib
csl: manuscript_extra/elsevier-harvard_rjt.csl
execute: 
  echo: false
  error: false
editor: 
  markdown: 
    wrap: sentence
---

@@

# tar_load("summarised_bird_singletons")
# tar_load("summarised_bird")
# tar_load("testate_summ")
 tar_load("last_chance_summ")
# tar_load("pollen_assem_res")
# tar_load("pollen_dataset_res")
 tar_load("pollen_results")
 tar_load("hershop_pollen")
 tar_load("wildhorse")
 tar_load("pollen3")
 tar_load("pollen1")
 tar_load("pollen2")
 tar_load("diatom1_summ")
 tar_load("diatom2_summ")
 tar_load("chironomid1_summ")
 tar_load("marine1_summ")
# tar_load("chironomid2_summ")

@@

Counts of organisms are a key data type for ecology and palaeoecology.
The sum of the counts is an important metric of data quality: larger count sums are associated with smaller uncertainties, both in relative abundance of taxa and any derived statistics such as transfer function reconstructions [@Heiri2001; @Payne2009].
For raw count data, the count sum can be easily calculated, but some of the considerable amount of archived count data are only available as percentages.
For some percent datasets, the count sum has not been archived and we might want to estimate it.

While it is, in general, not possible to calculate the denominator used to calculate percentages, this paper aims to show that it is possible to estimate it for community or assemblage count data.
Two insights make this possible.
The first is that the raw counts are typically integers, giving a granularity [@Brown2017] to the percentage data. The second is that the raw counts can be represented by a species abundance distribution. The ubiquitous shape of species abundance distributions when plotted as the number of species against abundance is a hyperbolic curve with many rare and few abundant species [@McGill_2007].
From this shape, which is central to ecological theory [@McGill_2007], two predictions can be made: that counts of an entire community or assemblage should have many rare taxa, typically including taxa occuring only once (hereafter singletons) and that the distribution of abundances should have a smooth and monotonic decline.

This paper will use these insights to develop methods for estimating the count sum from percent data, and demonstrate them with some percent data.
Then the paper tests the assumptions of the methods with a large collection of ecological and palaeoecological count data.
This reveals that a some datasets have count data that differ markedly from the predictions based on species abundance distributions.
These datasets might not accurately reflect the original community or assemblage.

<!-- Six percent of papers published in *Molecular and Cellular Biology* show evidence of inappropriate image duplication [@Bik2018] either due to error or, more rarely, misconduct. -->
<!-- It would, especially after the spate of retractions in behavioural ecology [@Pennisi2021], be reckless to assume that the ecological and palaeoecological literature is devoid of errors and misconduct. -->
<!-- The low rate of retractions in the ecological and geological literature [@Grieneisen2012] may partially reflect our limited ability to detect errors and misconduct rather than their prevalence. -->

<!-- Several numerical tools have been developed to identify questionable data. -->
<!-- Deviations from the distribution of digits expected from the Newcomb-Benford law have been used to detect issues with scientific and financial data [@Barabesi2017]. -->
<!-- @Carlisle2017 identified papers where the baseline differences in means for different treatments were surprisingly high or low given the variance of the data. -->
<!-- @Brown2017 developed the granularity-related inconsistency of means (GRIM) test which assesses whether means of integer data are consistent with the reported sample size. -->
<!-- However none of these tests are directly applicable to microfossil assemblage data, one of the most common types of palaeoecological data. -->

<!-- A common assertion in papers reporting microfossil assemblage data is that a minimum of *N* microfossils were counted, where *N* is often fifty for chironomids and several hundred for pollen and diatoms. -->

<!-- There is a risk that count sums will be misreported. -->
<!-- This could happen by mistake, especially with the use of secondary data. -->
<!-- It could also happen knowingly, especially when microfossil preservation is poor or concentrations are low. -->
<!-- This risk that count sums could be misreported is demonstrated by @Larocque-Tobler2016 who acknowledged that the chironomid count sums in @Larocque-Tobler2015 were not at least fifty head capsules as reported but as low as nineteen. -->
<!-- @Telford2019 reports some other cases where count sums may be much lower than reported. -->

<!-- Given that count sums have been misreported, the possibility that other aspects of count data may be misreported needs to be considered. -->

@@

```{r}
neotoma_date <- list.files("data/neotoma/", pattern = "sql$") |> 
  str_extract("\\d{4}-\\d{1,2}-\\d{1,2}") |> 
  ymd() |> 
  max() |> 
  format("%d %B %Y")
```

Publicly available datasets were downloaded from the Palaeodata Center, Neotoma, Pangaea, and other sources.
A range of ecological and palaeoecological data were sought to allow for differences in typical count sums and species richness, and possible differences in species abundance distributions between taxonomic groups.

Count data analysed include the 1) North American Breeding Bird survey [@bbsAssistant2], a very large, annual monitoring programme to track the status of North American bird populations.
The survey reports all birds seen in point counts along \~40 km survey routes at the height of the breeding season.
2) The Brown University planktic foraminiferal database [@Prell1999].
3) The MOLTEN diatom database of modern coastal diatom assemblages [@Clarke2006].
4) The North American testate amoeba training set [@AMESBURY2018].
5) Fossil pollen assemblage from the Neotoma database (`r neotoma_date` snapshot) [@neotoma_2018].

Counts with with small count sums (\< 50), that appeared to be percentages, or were documented as back-transformed from digitised data were excluded from all datasets.

@@

Data sets with a many samples without singletons, a greatest common divisor greater than one, or an unusually high proportion of counts divisible by $k$, relative to the majority of pollen data set in the Neotoma database, were flagged for closer inspection.

@@

In general, it is not possible to determine the sum from which percentages were calculated, but the properties of community and assemblage counts make it possible to estimate it.
Given the expectation that the rarest taxon is a singleton, the count sum *N* can be estimated as $1/p_{min} \times 100$ where *p~min~* is the percent abundance of the rarest taxon.
This method will fail for assemblages without singletons.
It is also expected that the percent *p* is calculated from integer counts, therefore there should be a count sum *N* such that $p_i/100\times N$ is, within rounding error, an integer for all *i* taxa.
Possible values for *N* can be found by a direct search over the range of plausible values of *N*.
An infinite number of possible count sums that are consistent with the percent exist, but the lowest will give the correct value of *N* except in cases where the greatest common divisor is greater than one.
These tests for percent data are closely related to GRIM [@Brown2017], as all rely on the granularity of percentages calculated from integer data.

@@

All analyses were done in R v. `r getRversion()` [@R] and used the packages extraDistr v. `r packageVersion("extraDistr")` [@extraDistr], numbers v. `r packageVersion("numbers")` [@numbers], countSum v. `r packageVersion("countSum")` [@countSum], bbsAssistant v. `r packageVersion("bbsAssistant")` [@bbsAssistant1], and tidyverse v. `r packageVersion("tidyverse")` [@tidyverse].
Packages targets v. `r packageVersion("targets")` [@targets], quarto v. `r packageVersion("quarto")` [@quarto] and renv v. `r packageVersion("renv")` [@renv] were used to ensure reproducibility.
Code to reproduce all the analyses is archived at <https://github.com/richardjtelford/count.check.ms> and [zenodo.org](zenodo.org).
All issues identified in the Neotoma database were reported to the maintainers.

@@

Some datasets with assemblages without singletons have few taxa.
For example, the `r wildhorse$n_assemblage` assemblage dataset from Wildhorse Lake [@Mehringer1985] includes only `r wildhorse$n_taxa_dataset` taxa (`r mean(pollen_summ2$n_taxa_dataset > wildhorse$n_taxa_dataset) * 100` % of Neotoma pollen datasets have more than this), and relatively large count sums (median `r wildhorse$median_count`).
While `r wildhorse$no_singletons * 100`% of assemblages in this dataset lack singletons, even more assemblage (`r wildhorse$no_zeros * 100`%) lack taxa with zero counts.
This is in contrast to most (`r pollen_summ$has_zeros * 100`%) Neotoma pollen assemblages which have zero counts.\
These patterns could arise either when there is a high sampling effort in a low diversity site, or when the archived data do not include the entire pollen assemblage, instead focusing on the subset of abundant taxa of most interest to the authors.

@@

`r Words(pollen1$n_gcd2)` of the `r pollen1$n_sample` assemblages in dataset pollen1 lack singletons and have a greatest common divisor of two.
These assemblages have between `r pollen1$min_taxa` and `r pollen1$max_taxa` taxa.
The probability of all *n* counts being divisible by *k* is $k^{-n}$, assuming the distribution of counts is locally uniform.
That is, for *k* = 2, odd counts are as common as even counts.
For an assemblage with `r pollen1$min_taxa` or `r pollen1$max_taxa` taxa, the probability of having all counts divisible by two by chance are 10^`r ceiling(- pollen1$min_taxa * log10(2))`^ and 10^`r ceiling(- pollen1$max_taxa * log10(2))`^ respectively.
These probabilities are not so small that it is unrealistic to expect to find assemblages like these in the Neotoma database, but to find so many in one dataset is unexpected.
Furthermore, the assumption made here is optimistic, in reality for *k* = 2, more odd counts are expected ones are expected than even ones in most datasets, as typical rank abundance curves have more singletons than doubletons and so on, which makes finding assemblages with a greatest common divisor above one even more unlikely.

Dataset pollen2 has `r pollen2$n_sample` assemblages, of which `r words(pollen2$n_gcd2)` have a greatest common divisor of two and a further `r words(pollen2$n_gcd3)` with a greatest common divisor of three, one of which has `r pollen2$max_taxa3` taxa.
The probability of having all counts divisible by three in such a taxonomically rich assemblage is 10\^`r ceiling(- pollen2$max_taxa3 * log10(3))`.
In addition to the assemblages with a greatest common divisor above one, this dataset includes several assemblages that have a higher proportion of counts divisible by two or three than is typical in assemblages from Neotoma (@fig-multiples-plot).
It is unclear what processes could generate such counts.

@@

## Unexpected percent data

```{r}
#| label: line169

@@

Some taxonomic groups have microfossils that often come in groups of attached individuals.
For example, each diatom cell has two valves which are the counting unit.
These valves usually separate during processing, but paired valves are often found, and some taxa such as *Aulacoseira* produce long chains of strongly bound valves.
This might make singletons slightly less likely.
The results from Owen's Lake suggest that this is not an important problem, as most assemblages have many singletons.

@@

# Acknowledgements {.unnumbered}

@@

# References {.unnumbered}
